Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-05-09 src/tex/ms.tex
implement comments
mostly introduction

% Abstract
\begin{abstract}
Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
Thermal stratification of runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.

We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
We determine survey strategies that maximize the diagnostic power of the obtained data, and we identify as key mission design drivers: 1. A follow-up campaign of planetary mass measurements and 2. The fraction of low-mass stars in the target sample.
Observational constraints on the runaway greenhouse transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
\end{abstract}

@@

Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
Due to the inherent detection biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
These bodies experience conditions that are in some aspects comparable to the environment of the inner solar system bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.

An example from the solar system for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
Both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F} because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Way2020}.

Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.

The runaway greenhouse transition is a robust prediction from climate models~\citep{Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is sensitive on the water content.
However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,Kopparapu2013}.
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
Its prevalent definition refers to the possibility of liquid water on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.

@@

We find that retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$:
Dilution factors $f_\mathrm{rgh} \gtrsim 0.5$ yield posterior probability distributions that are condensed at the order of magnitude of the true value; accurate constraints to within $\pm 0.25\,\mathrm{dex}$ of the truth are reached only from $f_\mathrm{rgh} \approx 0.8$.

@@

\subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}\ref{sec:dis_keyfactors}
What influences the probability of correctly rejecting a false null hypothesis?
We identified the following six drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
\begin{itemize}
    \item Occurrence rate of planets forming steam atmospheres
    \item Planetary evolution and duration of the steam atmosphere phase
    \item Prevalent water inventory
    \item Size and composition of the planetary sample
    \item Radius measurement precision
    \item Availability and precision of mass measurements
\end{itemize}
We will now briefly explore the above drivers.

@@

\textit{Size and composition of the planetary sample}: Evidently, the significance of a statistical trend increases with a larger sample size.
In addition, the sample must include planets on both sides of the instellation threshold.
Due to the strongly distance-dependent detection bias associated with the transit method and the temporal coverage of upcoming transit missions, this is only likely for low-mass host stars.

@@

\subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
Confirming or disproving the predicted habitable zone inner edge discontinuity will depend on the significance with which the null hypothesis can be excluded, which is a function of instrumentation and survey strategy.
As discussed in Section~\label{sec:dis_keyfactors}, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission.% (Matuszewski, priv. comm.).
%In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} for hundreds of planets~\citep{Rauer2021}, the \plato\ mission is comparable to the optimistic survey (see Section~\ref{sec:res_testability}) in terms of sample size and precision.
If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.
The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
Overall, it seems feasible to derive the typical water content of low-mass exoplanets from these occurrence estimates.

@@

\section{Conclusions}\label{sec:conclusions}
Significant inflation of rocky planet radii is a robust prediction of runaway greenhouse models.
Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius/density discontinuity resulting from this inflation in the exoplanet population.
Our key findings are as follows:
\begin{enumerate}
    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
    \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.
          For a planet sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to confirm or reject the predicted trend.
          Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements.
          This can reduce the required planet yield by about a factor of three.
          Only an adequate sample of planets orbiting M~dwarfs will ensure sufficient targets on both sides of the expected threshold instellation.
    \item Testing the runaway greenhouse hypothesis on a population level can provide constraints on the water inventory of rocky exoplanets and thus make an important contribution to assessing their habitability.
         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.

>>>>

% Abstract
\begin{abstract}
Long-term magma ocean phases on rocky exoplanets orbiting closer to their star than the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
Thermal stratification of runaway planets is expected to significantly inflate their atmospheres, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.

We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for sample sizes $\gtrsim$~100~planets if at least \SI{\sim 10}{\percent} of those orbiting closer than the habitable zone inner edge harbor runaway climates.
%We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
We determine survey strategies that maximize the diagnostic power of the obtained data and identify as key mission design drivers: 1. A follow-up campaign of planetary mass measurements and 2. The fraction of low-mass stars in the target sample.
Observational constraints on the runaway greenhouse transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
\end{abstract}

@@

Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
Due to the inherent biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
These experience thermal states that are in some aspects comparable to the ones of the inner solar system bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
Studying the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.

An example from the solar system for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K,Salvador2023b}.
While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
Both planets likely underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
Magma ocean states play a substantial role in establishing the long-term geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F}, in particular owing to efficient heat and volatile transfers between interior and atmosphere in the absence of a stiff boundary separating them~\citep{2021ApJ...909L..22K,Dorn2021,Salvador2023b}.
% because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus' water loss -- occurred late in their evolution~\citep[e.g.,][]{Way2020}.

Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}.
Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.

The runaway greenhouse transition is a robust prediction from climate models~\citep{Kasting1988,Nakajima1992,Goldblatt2012,Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
In particular, planets in a runaway greenhouse state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is a function of the water content.
However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and chemical exchange between core and mantle material may influence the amount and speciation of outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
Astronomical observations of planets that are currently in a runaway greenhouse state may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp instellation threshold~\citep{Goldblatt2013,Leconte2013,Kopparapu2013}.
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone~\citep[e.g.,][]{Ramirez2018b,Salvador2023b}.
Its prevalent definition refers to the possibility of a sustained liquid water body on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014};
its exact spatial location and extent may be strongly influenced by the interior and atmosphere oxidation state and resulting atmosphere composition~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity remains controversial~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries defined by stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.

@@

We find that retrievals from radius measurements alone require high fractions of greenhouse climate-bearing planets to achieve an accurate constraint on $S_\mathrm{thresh}$:
dilution factors $f_\mathrm{rgh} \gtrsim 0.5$ yield posterior probability distributions that are condensed at the order of magnitude of the true value; accurate constraints to within $\pm 0.25\,\mathrm{dex}$ of the truth are reached only from $f_\mathrm{rgh} \approx 0.8$.

@@

\subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}\label{sec:dis_keyfactors}
What influences the probability of correctly rejecting a false null hypothesis?
We identified the following six drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
\begin{itemize}
    \item Occurrence rate of planets forming steam atmospheres
    \item Planetary evolution and duration of the steam atmosphere phase
    \item Prevalent water inventory
    \item Size and composition of the planetary sample
    \item Radius measurement precision
    \item Availability and precision of mass measurements
\end{itemize}
We will now briefly explore the above drivers.

@@

\textit{Size and composition of the planetary sample}: The significance of a statistical trend increases with a larger sample size.
In addition, the sample must include planets on both sides of the instellation threshold.
Due to the strongly distance-dependent detection bias associated with the transit method and the temporal coverage of upcoming transit missions, this is only likely for low-mass host stars.

@@

\subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
Confirming or disproving the predicted habitable zone inner edge discontinuity will depend on the significance with which the null hypothesis can be excluded, which is a function of instrumentation and survey strategy.
As discussed in Section~\ref{sec:dis_keyfactors}, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission.% (Matuszewski, priv. comm.).
%In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} for hundreds of planets~\citep{Rauer2021}, the \plato\ mission is comparable to the optimistic survey (see Section~\ref{sec:res_testability}) in terms of sample size and precision.
If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.
The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
Overall, it seems feasible to derive the typical water content of low-mass exoplanets from these occurrence estimates.

@@

\section{Conclusions}\label{sec:conclusions}
Significant inflation of rocky planet radii is a robust prediction of runaway greenhouse models.
Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius/density discontinuity resulting from this inflation in the exoplanet population.
Our key findings are as follows:
\begin{enumerate}
    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to their irradiation.
    \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.
          For a planet sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to confirm or reject the predicted trend.
          Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements.
          This can reduce the required planet yield by about a factor of three.
          Only an adequate sample of planets orbiting M~dwarfs will ensure sufficient targets on both sides of the expected threshold instellation.
    \item Testing the runaway greenhouse hypothesis on a population level can provide constraints on the water inventory of rocky exoplanets and thus make an important contribution to assessing their habitability.
         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.
